Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-22 appendix-EPQ.tex
modify

Let $\textit{Items}(e,p)$ be the set of items with priority $p$ in execution $e$. The following lemma states a method to build the linearization of $e$ w.r.t $\mathsf{MatchedMaxPriority}^{=}$.

@@

\item[-] For item $z$ with priority smaller than $p$. If both $\textit{call}(\textit{put},z,\_)$ and $\textit{call}(\textit{rm},z)$ is before $o$, then locate the linearization points of $\textit{put}(z,\_)$ and $\textit{rm}(z)$ just after their call actions. If both $\textit{ret}(\textit{put},z)$ and $\textit{ret}(\textit{rm},z)$ (if exists) is after $o$, then locate the linearization points of $\textit{put}(z,\_)$ and $\textit{rm}(z)$ just before their return actions. Otherwise, $x$ is in interval of $z$, which contradicts the definition of gap-point, and is impossible.
\end{itemize}

Let $l$ be the sequence of linearization points constructed above. It is obvious that $e \sqsubseteq l$. Since for each $y \in \textit{Items}(e,p)$, $o$ is after $\textit{call}(\textit{put},y,\_)$ and $\textit{call}(\textit{rm},x)$, we can see that $\textit{rm}(x)$ is after $\textit{put}(y,p)$ and $\textit{rm}(y)$ in $l$. It is obvious that $\textit{put}(x,p)$ is before $\textit{rm}(x)$ in $l$. Since $x$ does not $<_{\textit{pb}}$ to $y$, we can see that no $\textit{put}(y,p)$ happens before $\textit{put}(x,p)$. Then it is easy to see that $\textit{put}(x,p)$ is after $\textit{put}(y,p)$ in $l$. Since $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e)$ holds, all other items in $\textit{Items}(e,p)$ has matched $\textit{put}$ and $\textit{rm}$, and it is easy to see that their $\textit{put}$ and $\textit{rm}$ (except for that of $x$) are all before $\textit{rm}(x)$ in $l$.

For item $z$ with priority smaller than $p$, we can see that there are only two possibilities: (1) $\textit{put}(z,\_)$ and $\textit{rm}(z)$ are both before $\textit{rm}(x)$ in $l$, and (2) $\textit{put}(z,\_)$ and $\textit{rm}(z)$ (if exists) are after before $\textit{rm}(x)$ in $l$. Therefore, before $\textit{rm}(x)$ in $l$, the $\textit{put}$ and $\textit{rm}$ of $z$ are matched.

@@

According to the definition of $<_{\textit{ob}}^*$, if $a <_{\textit{pb}}^* b$, then there exists $a_1,\ldots,a_m$, such that $a <_{\textit{pb}} a_1 <_{\textit{pb}} \ldots <_{\textit{pb}} a_m <_{\textit{pb}} b$. The following lemma states that, the number of intermediate items $a_i$ is in fact bounded.

@@

\begin{restatable}{lemma}{FiveEnmuerationisEnoughForEPQOneEqual}
\label{lemma:five enumeration is enough for EPQ1Equal}
Given a data-differentiated $p$-execution $e$ where $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e)$ holds. Let $a$ and $b$ be items with maximal priority $p$. Assume that $a <_{\textit{pb}}^* b$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$. Then, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists), where $a_1$ is the possible intermediate value for obtain $a <_{\textit{pb}}^* b$.
\end{restatable}
\begin {proof}

@@

Given a data-differentiated execution $e$ with only one maximal priority, two actions $\textit{act}_1$, $\textit{act}_2$ of maximal priority in $e$, and assume that $\textit{act}_1$ is before $\textit{act}_2$ in $e$.
we say that $\textit{act}_1$, $\textit{act}_2$ is covered by items $d_1,\ldots,d_m$ in $e$, if the priorities of $d_1,\ldots,d_m$ is smaller than that of $\textit{act}_1$ and $\textit{act}_2$, and

@@

Let us begin to represent several register automata that is used to capture the existence of a data-differentiated execution $e$, $e$ has a projection $e'$, $e'$ has only one maximal priority, $\mathsf{Has\text{-}MatchedMax-}$ $\mathsf{Priority}(e')$ holds, there exists items $a$ and $b$ with maximal priority in $e'$, $a <_{\textit{pb}}^* b$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,\_)$ or $\textit{call}(\textit{rm},a)$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\textit{1-eq}} \cap \mathcal{I} \neq \emptyset$ if and only if there exists $e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$, $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds, $x$ is the value with maximal priority in $e'$, $e'$ has only one maximal priority, $a$ and $b$ are two items with maximal priority $p$ in $e'$, $a <_{\textit{pb}}^* b$ in $e'$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e'$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by some register automata in $\mathcal{A}_{\textit{1-eq}}$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Since $e_1$ is accepted by some register automata in  $\mathcal{A}_{\textit{1-eq}}$, let $x$, $y$ and $z$ (if exists) be the items that are renamed into $b$, $a$ and $a_1$ (if exists) by $r$, respectively, and let $d_1,\ldots,d_m$ be the items that are renamed into $d$ by $r$.

@@

\noindent The $\textit{if}$ direction: Assume that there exists $e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds $e'$ has only one maximal priority, $a'$ and $b'$ are two items with maximal priority $p$ in $e'$, $a' <_{\textit{pb}}^* b'$ in $e'$, and the rightmost gap-point of $b'$ is before $\textit{call}(\textit{put},a',p)$ or $\textit{call}(\textit{rm},a')$ in $e'$. By data-independence, we can obtain execution $e_1$ as follows: (1) rename $a'$ and $b'$ into $a$ and $b$, respectively, (2) for the items $d_1,\ldots,d_m$ that constitute the rightmost gap of $b'$, we rename them into $d$, (3) if $a' <_{\textit{pb}}^A a'_1 <_{\textit{pb}}^B b$, we rename $a'_1$ into $a_1$, and (4) rename the other items into $\top$. It is easy to see that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e_1)$ holds, $a$ and $b$ has maximal priority in $e_1$, $a <_{\textit{pb}}^* b$ in $e_1$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e_1$. By Lemma \ref{lemma:five enumeration is enough for EPQ1Equal}, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists). Then

@@

Given a data-differentiated execution $e$, we say that $o = \textit{rm}(\textit{empty})$ in $e$ is covered by items $d_1,\ldots,d_m$, if

@@

According to the definition of left-right constraint for $o$, in a data-differentiated execution $e$, there is a cycle going through $o$, if and only if there exists items $d_1,\ldots,d_m$, such that $o$ is covered by $d_1,\ldots,d_m$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\seqPQ}^3 \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, $\mathsf{Has\text{-}EmptyRemove}(e')$ holds, $o = \textit{rm}(\textit{empty})$ is in $e'$, and $o$ is covered by some items $d_1,\ldots,d_m$ in $e'$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by $\mathcal{A}_{\seqPQ}^3$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Let $d_1,\ldots,d_m$ be the items in $e_2$ such that $r(d_i)=b$ for each $1 \leq i \leq m$. Let $e_3 = e_2 \vert_{ \{ o, d_1, \ldots, d_m \} }$. It is obvious that $e_3 \in \textit{proj}(e_2)$ and $\textit{last}(e_3) = \seqPQ_3$. It is easy to see that $o$ is covered by $d_1,\ldots,d_m$.

\noindent The $\textit{if}$ direction: Assume that there exists such $e$, $e'$, $o$ and $d_1,\ldots,d_m$. Then, let $e_1$ be obtained from $e$ by renaming $d_1,\ldots,d_m$ into $b$ and renaming other items into $\top$. By data-independence, $e_1 \in \mathcal{I}$. It is easy to see that $e_1$ is accepted by $\mathcal{A}_{\seqPQ}^3$.

>>>>

Let $\textit{Items}(e,p)$ be the set of values with priority $p$ in execution $e$. The following lemma states a method to build the linearization of $e$ w.r.t $\mathsf{MatchedMaxPriority}^{=}$.

@@

\item[-] For value $z$ with priority smaller than $p$. If both $\textit{call}(\textit{put},z,\_)$ and $\textit{call}(\textit{rm},z)$ is before $o$, then locate the linearization points of $\textit{put}(z,\_)$ and $\textit{rm}(z)$ just after their call actions. If both $\textit{ret}(\textit{put},z)$ and $\textit{ret}(\textit{rm},z)$ (if exists) is after $o$, then locate the linearization points of $\textit{put}(z,\_)$ and $\textit{rm}(z)$ just before their return actions. Otherwise, $x$ is in interval of $z$, which contradicts the definition of gap-point, and is impossible.
\end{itemize}

Let $l$ be the sequence of linearization points constructed above. It is obvious that $e \sqsubseteq l$. Since for each $y \in \textit{Items}(e,p)$, $o$ is after $\textit{call}(\textit{put},y,\_)$ and $\textit{call}(\textit{rm},x)$, we can see that $\textit{rm}(x)$ is after $\textit{put}(y,p)$ and $\textit{rm}(y)$ in $l$. It is obvious that $\textit{put}(x,p)$ is before $\textit{rm}(x)$ in $l$. Since $x$ does not $<_{\textit{pb}}$ to $y$, we can see that no $\textit{put}(y,p)$ happens before $\textit{put}(x,p)$. Then it is easy to see that $\textit{put}(x,p)$ is after $\textit{put}(y,p)$ in $l$. Since $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e)$ holds, all other values in $\textit{Items}(e,p)$ has matched $\textit{put}$ and $\textit{rm}$, and it is easy to see that their $\textit{put}$ and $\textit{rm}$ (except for that of $x$) are all before $\textit{rm}(x)$ in $l$.

For value $z$ with priority smaller than $p$, we can see that there are only two possibilities: (1) $\textit{put}(z,\_)$ and $\textit{rm}(z)$ are both before $\textit{rm}(x)$ in $l$, and (2) $\textit{put}(z,\_)$ and $\textit{rm}(z)$ (if exists) are after before $\textit{rm}(x)$ in $l$. Therefore, before $\textit{rm}(x)$ in $l$, the $\textit{put}$ and $\textit{rm}$ of $z$ are matched.

@@

According to the definition of $<_{\textit{ob}}^*$, if $a <_{\textit{pb}}^* b$, then there exists $a_1,\ldots,a_m$, such that $a <_{\textit{pb}} a_1 <_{\textit{pb}} \ldots <_{\textit{pb}} a_m <_{\textit{pb}} b$. The following lemma states that, the number of intermediate values $a_i$ is in fact bounded.

@@

\begin{restatable}{lemma}{FiveEnmuerationisEnoughForEPQOneEqual}
\label{lemma:five enumeration is enough for EPQ1Equal}
Given a data-differentiated $p$-execution $e$ where $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e)$ holds. Let $a$ and $b$ be values with maximal priority $p$. Assume that $a <_{\textit{pb}}^* b$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$. Then, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists), where $a_1$ is the possible intermediate value for obtain $a <_{\textit{pb}}^* b$.
\end{restatable}
\begin {proof}

@@

Given a data-differentiated execution $e$ with only one maximal priority, two actions $\textit{act}_1$, $\textit{act}_2$ of maximal priority in $e$, and assume that $\textit{act}_1$ is before $\textit{act}_2$ in $e$.
we say that $\textit{act}_1$, $\textit{act}_2$ is covered by values $d_1,\ldots,d_m$ in $e$, if the priorities of $d_1,\ldots,d_m$ is smaller than that of $\textit{act}_1$ and $\textit{act}_2$, and

@@

Let us begin to represent several register automata that is used to capture the existence of a data-differentiated execution $e$, $e$ has a projection $e'$, $e'$ has only one maximal priority, $\mathsf{Has\text{-}MatchedMax-}$ $\mathsf{Priority}(e')$ holds, there exists values $a$ and $b$ with maximal priority in $e'$, $a <_{\textit{pb}}^* b$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,\_)$ or $\textit{call}(\textit{rm},a)$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\textit{1-eq}} \cap \mathcal{I} \neq \emptyset$ if and only if there exists $e \in \mathcal{I}$ and $e' \in \textit{proj}(e)$, $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds, $x$ is the value with maximal priority in $e'$, $e'$ has only one maximal priority, $a$ and $b$ are two values with maximal priority $p$ in $e'$, $a <_{\textit{pb}}^* b$ in $e'$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e'$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by some register automata in $\mathcal{A}_{\textit{1-eq}}$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Since $e_1$ is accepted by some register automata in  $\mathcal{A}_{\textit{1-eq}}$, let $x$, $y$ and $z$ (if exists) be the values that are renamed into $b$, $a$ and $a_1$ (if exists) by $r$, respectively, and let $d_1,\ldots,d_m$ be the values that are renamed into $d$ by $r$.

@@

\noindent The $\textit{if}$ direction: Assume that there exists $e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, such that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e')$ holds $e'$ has only one maximal priority, $a'$ and $b'$ are two values with maximal priority $p$ in $e'$, $a' <_{\textit{pb}}^* b'$ in $e'$, and the rightmost gap-point of $b'$ is before $\textit{call}(\textit{put},a',p)$ or $\textit{call}(\textit{rm},a')$ in $e'$. By data-independence, we can obtain execution $e_1$ as follows: (1) rename $a'$ and $b'$ into $a$ and $b$, respectively, (2) for the values $d_1,\ldots,d_m$ that constitute the rightmost gap of $b'$, we rename them into $d$, (3) if $a' <_{\textit{pb}}^A a'_1 <_{\textit{pb}}^B b$, we rename $a'_1$ into $a_1$, and (4) rename the other values into $\top$. It is easy to see that $\mathsf{Has\text{-}MatchedMaxPriority}^{=}(e_1)$ holds, $a$ and $b$ has maximal priority in $e_1$, $a <_{\textit{pb}}^* b$ in $e_1$, and the rightmost gap-point of $b$ is before $\textit{call}(\textit{put},a,p)$ or $\textit{call}(\textit{rm},a)$ in $e_1$. By Lemma \ref{lemma:five enumeration is enough for EPQ1Equal}, there are five possible enumeration of operations of $a$, $b$, $a_1$ (if exists). Then

@@

Given a data-differentiated execution $e$, we say that $o = \textit{rm}(\textit{empty})$ in $e$ is covered by values $d_1,\ldots,d_m$, if

@@

According to the definition of left-right constraint for $o$, in a data-differentiated execution $e$, there is a cycle going through $o$, if and only if there exists values $d_1,\ldots,d_m$, such that $o$ is covered by $d_1,\ldots,d_m$.

@@

\noindent {\bf $\textit{fact}_1$}: Given a data-independent implementation $\mathcal{I}$. $\mathcal{A}_{\seqPQ}^3 \cap \mathcal{I} \neq \emptyset$ if and only if $\exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e)$, $\mathsf{Has\text{-}EmptyRemove}(e')$ holds, $o = \textit{rm}(\textit{empty})$ is in $e'$, and $o$ is covered by some values $d_1,\ldots,d_m$ in $e'$.

\noindent The $\textit{only if}$ direction: Assume that $e_1 \in \mathcal{I}$ is accepted by $\mathcal{A}_{\seqPQ}^3$. By data-independence, there exists data-differentiated execution $e_2 \in \mathcal{I}$ and a renaming function $r$, such that $e_1=r(e_2)$. Let $d_1,\ldots,d_m$ be the values in $e_2$ such that $r(d_i)=b$ for each $1 \leq i \leq m$. Let $e_3 = e_2 \vert_{ \{ o, d_1, \ldots, d_m \} }$. It is obvious that $e_3 \in \textit{proj}(e_2)$ and $\textit{last}(e_3) = \seqPQ_3$. It is easy to see that $o$ is covered by $d_1,\ldots,d_m$.

\noindent The $\textit{if}$ direction: Assume that there exists such $e$, $e'$, $o$ and $d_1,\ldots,d_m$. Then, let $e_1$ be obtained from $e$ by renaming $d_1,\ldots,d_m$ into $b$ and renaming other values into $\top$. By data-independence, $e_1 \in \mathcal{I}$. It is easy to see that $e_1$ is accepted by $\mathcal{A}_{\seqPQ}^3$.
